---
title: "Reproducible Research CP1"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---
##Loading required R packages
```{r}
library(knitr)
library(dplyr)
library(lubridate)
library(ggplot2)
Sys.setlocale("LC_TIME", "C")
```

## Procject Description

###The purpose of this project was to practice:

*loading and preprocessing data
*imputing missing values
*interpreting data to answer research questions

##Data Description:

###The data for this assignment was downloaded from the course web site:

Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

*steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

*date: The date on which the measurement was taken in YYYY-MM-DD format

*interval: Identifier for the 5-minute interval in which measurement was taken

*The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.*

##Reading the data

```{r}
data <- read.csv("activity.csv",sep = ",")
```

##What is mean total number of steps taken per day?

*For this part of the assignment the missing values can be ignored.*

1.Calculate the total number of steps taken per day.
2.Make a histogram of the total number of steps taken each day.
3.Calculate and report the mean and median of the total number of steps taken per day.

###Progress:

1.Calculate the total number of steps taken per day.
```{r}
steps <- data %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```
2.Making the histogram:
```{r}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "black", binwidth = 1000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")
```
3.Calculate the mean and median of the total number of steps taken per day:
```{r}
mean_steps <- mean(steps$steps, na.rm = TRUE)
median_steps <- median(steps$steps, na.rm = TRUE)
```

```{r}
mean_steps
```


```{r}

median_steps
```

##What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?.

###Progress:
*Calculate the average number of steps taken in each 5-minute interval per day and group by interval:*

```{r}
interval <- data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))
```

1.Making the time series of the 5-minute interval and average steps:

```{r}
ggplot(interval, aes(x=interval, y=steps)) +
  geom_line(color = "black")
```

2.Find out the maximum steps, on average, across all the days:

```{r}
interval[which.max(interval$steps),]
```

##Imputing missing values

*Note that there are a number of days/intervals where there are missing values (NA). The presence of missing days may introduce bias into some calculations or summaries of the data.*

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

4.Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

###Progress

1.Summarize all the missing values:

```{r}
sum(is.na(data$steps))
```

Missing values are 2304.

2. Filling in a missing NA with the average number of steps in the same 5-min interval.

3.Creating a new dataset as the original and use tapply for filling in the missing values with the average number of steps per 5-minute interval:

```{r}
data_full <- data
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm=TRUE, simplify=TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]
```

Check that there are no missing values:

```{r}
sum(is.na(data_full$steps))
```


No missing values.

4. Calculate the number of steps taken in each 5-minute interval per day using dplyr and group by interval. Use ggplot for making the histogram:

```{r}
steps_full <- data_full %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

```{r}
ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "black", binwidth = 1000) +
  labs(title = "Histogram of Steps per day, including missing values", x = "Steps per day", y = "Frequency")
```


###Calculate the mean and median steps with the filled in values:

```{r}
mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)
median_steps_full <- median(steps_full$steps, na.rm = TRUE)
```

```{r}
mean_steps_full
```

```{r}
median_steps_full
```
The impact of imputing missing data with the average number of steps in the same 5-min interval is that both the mean and the median are equal to the same value: 10766.

##Are there differences in activity patterns between weekdays and weekends?

*For this part the weekdays() will come handy. Use the dataset with the filled-in missing values for this part.*

1. Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
###Progress:
1. Creating new column, weektype, and apply whether the day is weekend or weekday:

```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(as.POSIXlt(data_full$date)) == "Saturday" | weekdays(as.POSIXlt(data_full$date)) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)

```

2. Calculate the average steps in the 5-minute interval and use ggplot for making the time series of the 5-minute interval for weekday and weekend, and compare the average steps:

```{r}
interval_full <- data_full %>% group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
s<- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```

From the two plots it seems that the test object is more active earlier in the day during weekdays compared to weekends, but more active throughout the weekends compared with weekdays (probably because the oject is working during the weekdays, hence moving less during the day).